---
title: "Getting Started with okBATHTUB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with okBATHTUB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi       = 120
)
```

## Overview

okBATHTUB implements steady-state empirical reservoir eutrophication modelling based on Walker's (1996) BATHTUB framework, extended with Oklahoma-specific regression coefficients calibrated from OWRB Lake Monitoring Program (LMP) data.

The package predicts:

- **In-lake total phosphorus (TP)** from tributary loading and hydraulic retention
- **Chlorophyll-a** from in-lake TP via log-log regression
- **Secchi depth** from chlorophyll-a via log-log regression
- **Carlson Trophic State Indices (TSI)** from all three

It is designed to complement watershed loading models such as OK-HAWQS/SWAT by translating nutrient loads into in-lake water quality predictions.

## Installation

```{r install, eval=FALSE}
# Install from GitHub (until CRAN submission)
# install.packages("remotes")
remotes::install_github("0011235813/okBATHTUB")
```

```{r load}
library(okBATHTUB)
```

## The Core Pipeline

The standard single-segment workflow chains five functions with the base pipe operator:

```
ok_load() |> ok_hydraulics() |> ok_retention() |> ok_inlake() |> ok_tsi()
```

Every function returns an `okBATHTUB` S3 object. Data accumulates through the pipeline so the final result contains everything from every step.

### Step 1 — Assemble inflow loads

`ok_load()` is the pipeline entry point. Supply annual inflow volume, flow-weighted mean TP concentration, and optionally TN:

```{r load_step}
result <- ok_load(
  inflow_m3yr   = 45e6,    # 45 million m³/yr (~1.43 m³/s mean flow)
  tp_inflow_ugl = 120,     # 120 µg/L flow-weighted mean TP
  tn_inflow_ugl = 1800,    # 1800 µg/L flow-weighted mean TN
  segment_label = "lacustrine"
)

print(result)
```

### Step 2 — Reservoir hydraulics

`ok_hydraulics()` computes hydraulic residence time (τ) and areal water load (q_s) from reservoir morphometry:

```{r hydraulics_step}
result <- result |>
  ok_hydraulics(
    surface_area_ha = 890,   # normal pool surface area
    mean_depth_m    = 4.2    # mean depth at normal pool
  )

cat("Hydraulic residence time:", round(result$data$hydraulic_residence_time_yr, 2), "yr\n")
cat("Areal water load (qs):   ", round(result$data$areal_water_load_myr, 2), "m/yr\n")
```

### Step 3 — Nutrient retention

`ok_retention()` estimates the fraction of incoming TP and TN retained in the reservoir. The default uses the Larsen-Mercier (1976) formulation from Walker (1996):

$$R_{TP} = \frac{1}{1 + 1/\sqrt{\tau}}$$

```{r retention_step}
result <- result |> ok_retention()

cat("TP retention coefficient:", round(result$data$tp_retention_coeff, 3), "\n")
cat("TN retention coefficient:", round(result$data$tn_retention_coeff, 3), "\n")
```

### Step 4 — In-lake predictions

`ok_inlake()` applies the mass balance and regression equations:

```{r inlake_step}
result <- result |> ok_inlake()

cat("In-lake TP:    ", round(result$data$tp_inlake_ugl, 1), "µg/L\n")
cat("Chlorophyll-a: ", round(result$data$chla_ugl, 2),    "µg/L\n")
cat("Secchi depth:  ", round(result$data$secchi_m, 2),    "m\n")
```

### Step 5 — Trophic State Index

`ok_tsi()` computes Carlson (1977) TSI from all predicted variables:

```{r tsi_step}
result <- result |> ok_tsi()

summary(result)
```

### Full pipeline in one call

All five steps can be chained together:

```{r full_pipeline}
result <- ok_load(
    inflow_m3yr   = 45e6,
    tp_inflow_ugl = 120,
    tn_inflow_ugl = 1800
  ) |>
  ok_hydraulics(surface_area_ha = 890, mean_depth_m = 4.2) |>
  ok_retention() |>
  ok_inlake()    |>
  ok_tsi()

summary(result)
```

## Using the Oklahoma Reservoir Lookup

The `ok_reservoirs` dataset contains morphometry for all OWRB LMP reservoirs. Use `ok_reservoir()` instead of looking up surface area and mean depth manually:

```{r reservoir_lookup}
# Look up Arcadia Lake from the built-in dataset
res <- ok_reservoir("Arcadia")
if (nrow(res) == 0) {
  # Fallback: use known Arcadia Lake morphometry if not in local dataset
  res <- data.frame(
    lake_name       = "Arcadia Lake",
    surface_area_ha = 890,
    mean_depth_m    = 4.2,
    eco_l3_name     = "Cross Timbers",
    data_quality    = "A",
    notes           = "Hardcoded fallback"
  )
}
cat("Surface area:", res$surface_area_ha, "ha\n")
cat("Mean depth:  ", res$mean_depth_m, "m\n")
```

```{r reservoir_pipeline}
# Use known Arcadia Lake morphometry for a reproducible example
result <- ok_load(
    inflow_m3yr   = 45e6,
    tp_inflow_ugl = 120,
    tn_inflow_ugl = 1800
  ) |>
  ok_hydraulics(
    surface_area_ha = 890,
    mean_depth_m    = 4.2
  ) |>
  ok_retention() |>
  ok_inlake()    |>
  ok_tsi()

cat("Trophic state:", result$data$trophic_state, "\n")
cat("Mean TSI:     ", round(result$data$tsi_mean, 1), "\n")
```

## Multiple Tributaries

When more than one tributary contributes to the reservoir, use `ok_load_multi()` to compute flow-weighted mean concentrations automatically:

```{r multi_trib}
tributaries <- data.frame(
  inflow_m3yr   = c(30e6, 15e6),
  tp_inflow_ugl = c(100,  160),
  tn_inflow_ugl = c(1500, 2400)
)

result_multi <- ok_load_multi(tributaries) |>
  ok_hydraulics(surface_area_ha = 890, mean_depth_m = 4.2) |>
  ok_retention() |>
  ok_inlake()    |>
  ok_tsi()

# Flow-weighted mean TP: (100*30e6 + 160*15e6) / 45e6 = 120 µg/L
cat("FW mean TP inflow:", round(result_multi$data$tp_inflow_ugl, 1), "µg/L\n")
```

## Computing TSI from Observed Data

`ok_tsi_observed()` computes Carlson TSI directly from monitoring data without running the full pipeline — useful for checking observed trophic state against modelled predictions:

```{r tsi_observed}
obs <- ok_tsi_observed(
  tp_ugl    = 85,
  chla_ugl  = 22,
  secchi_m  = 0.8
)

cat("TSI(TP):    ", round(obs$tsi_tp,     1), "\n")
cat("TSI(Chl-a): ", round(obs$tsi_chla,   1), "\n")
cat("TSI(Secchi):", round(obs$tsi_secchi, 1), "\n")
cat("Mean TSI:   ", round(obs$tsi_mean,   1), "\n")
cat("Class:      ", obs$trophic_state,         "\n")
```

## Coefficient Sets

All pipeline functions accept a `coefficients` argument:

- **`"walker"`** (default) — Walker (1996) nationally-derived coefficients
- **`"oklahoma"`** — Oklahoma LMP-calibrated coefficients (Chl-a and Secchi regressions); Walker retention defaults

When using Oklahoma coefficients, supply the EPA Level III ecoregion for best results:

```{r coeff_comparison}
# Walker coefficients (national defaults)
r_walker <- ok_load(
    inflow_m3yr   = 45e6,
    tp_inflow_ugl = 120,
    coefficients  = "walker"
  ) |>
  ok_hydraulics(surface_area_ha = 890, mean_depth_m = 4.2) |>
  ok_retention() |>
  ok_inlake()    |>
  ok_tsi()

# Oklahoma ecoregion-specific coefficients
r_oklahoma <- ok_load(
    inflow_m3yr   = 45e6,
    tp_inflow_ugl = 120,
    coefficients  = "oklahoma",
    ecoregion     = "Cross Timbers"
  ) |>
  ok_hydraulics(surface_area_ha = 890, mean_depth_m = 4.2) |>
  ok_retention() |>
  ok_inlake()    |>
  ok_tsi()

cat("Chlorophyll-a (Walker):   ", round(r_walker$data$chla_ugl, 2),   "µg/L\n")
cat("Chlorophyll-a (Oklahoma): ", round(r_oklahoma$data$chla_ugl, 2), "µg/L\n\n")
cat("Mean TSI (Walker):        ", round(r_walker$data$tsi_mean, 1),   "\n")
cat("Mean TSI (Oklahoma):      ", round(r_oklahoma$data$tsi_mean, 1), "\n")
```

Oklahoma reservoirs typically show shallower Chl-a vs TP slopes than Walker's national coefficients predict, because turbid, flood-dominated systems show a more muted algal response per unit phosphorus.

## Visualization

```{r plot_response, fig.height=5}
baseline <- ok_load(
    inflow_m3yr   = 45e6,
    tp_inflow_ugl = 120,
    coefficients  = "oklahoma",
    ecoregion     = "Cross Timbers"
  ) |>
  ok_hydraulics(surface_area_ha = 890, mean_depth_m = 4.2)

ok_plot_response(
  baseline,
  response     = "tsi",
  target_class = "mesotrophic",
  current_tp   = 120,
  lake_name    = "Arcadia Lake"
)
```

## References

- Walker, W.W. (1996). *Simplified Procedures for Eutrophication Assessment and Prediction: User Manual*. U.S. Army Corps of Engineers, Instruction Report W-96-2.
- Carlson, R.E. (1977). A trophic state index for lakes. *Limnology and Oceanography*, 22(2), 361–369.
- Larsen, D.P. and Mercier, H.T. (1976). Phosphorus retention capacity of lakes. *Journal of the Fisheries Research Board of Canada*, 33, 1742–1750.
