---
title: "Linking OK-HAWQS/SWAT Outputs to okBATHTUB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linking OK-HAWQS/SWAT Outputs to okBATHTUB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi       = 120
)
```

## Overview

OK-HAWQS/SWAT and okBATHTUB are complementary models designed to work together in a two-model nutrient management workflow:

- **OK-HAWQS/SWAT** simulates watershed hydrology and nutrient loading — translating land management decisions into tributary flow, sediment, and nutrient yields at the watershed outlet
- **okBATHTUB** simulates in-lake response — translating those tributary loads into predicted reservoir water quality

Neither model alone spans the full management question. Together they answer: *"If we implement BMPs that reduce watershed TP export by X%, what will the reservoir's chlorophyll-a and trophic state be?"*

```{r load}
library(okBATHTUB)
library(dplyr)
library(readr)
```

## The Two-Model Workflow

```
Watershed                           Reservoir
─────────────────────────────────── ──────────────────────────────────
Land use / management inputs        Tributary loads from HAWQS
        ↓                                   ↓
   OK-HAWQS/SWAT                    ok_load() ──→ ok_hydraulics()
        ↓                                           ↓
Daily streamflow + loads            ok_retention() ──→ ok_inlake()
        ↓                                               ↓
Annual/seasonal load summaries      ok_tsi() ──→ ok_plot_response()
        ↓                                           ↓
  [Handoff point]              In-lake TP, Chl-a, Secchi, TSI
```

The handoff occurs at the watershed outlet / reservoir inlet. OK-HAWQS produces annual or seasonal mean tributary concentrations and flow volumes; okBATHTUB takes these as inputs.

## OK-HAWQS Output File Structure

OK-HAWQS/SWAT produces output files in the `TxtInOut` directory. The relevant files for reservoir modelling are:

| File | Contents | okBATHTUB use |
|---|---|---|
| `output.rch` | Streamflow and constituent loads at each reach | Inflow volume and TP/TN loads |
| `output.sub` | Subbasin water balance | Watershed area confirmation |
| `output.hru` | HRU-level outputs | BMP scenario comparison |

The `output.rch` file uses fixed-width format with columns including `FLOW_OUTcms` (flow in m³/s), `NO3_OUT` (nitrate), `ORGN_OUT` (organic N), `SOLP_OUT` (soluble P), and `ORGP_OUT` (organic P).

## Reading OK-HAWQS Output

```{r read_hawqs, eval=FALSE}
# Read SWAT output.rch file
# (Adjust column positions for your SWAT version)
read_swat_rch <- function(rch_path) {
  # Read fixed-width format
  raw <- read_fwf(
    rch_path,
    fwf_cols(
      rch_id    = c(1, 4),
      year      = c(5, 9),
      month     = c(10, 12),
      day       = c(13, 16),
      area_km2  = c(17, 26),
      flow_cms  = c(27, 38),   # FLOW_OUTcms
      no3_kgha  = c(83, 94),   # NO3_OUT
      orgn_kgha = c(95, 106),  # ORGN_OUT
      solp_kgha = c(107, 118), # SOLP_OUT
      orgp_kgha = c(119, 130)  # ORGP_OUT
    ),
    skip = 9,
    col_types = "iiiiidddddd"
  )
  raw
}
```

## Simulated Example: 604(b) Sensitive Water Supply Workflow

This example replicates the analytical approach used in the OWRB 604(b) BIL-funded watershed modeling project. We use simulated OK-HAWQS output representing annual TP and flow loads at the reservoir inlet for a Sensitive Water Supply reservoir.

```{r simulate_hawqs}
# Simulated OK-HAWQS annual output at reservoir inlet
# Replace with actual output.rch data in production use
hawqs_annual <- data.frame(
  year         = 2010:2022,
  flow_m3yr    = c(38.2, 52.1, 29.4, 44.7, 61.3, 35.8, 41.9,
                   48.2, 33.6, 57.4, 42.1, 36.8, 49.3) * 1e6,
  tp_load_kgyr = c(4584, 7314, 2823, 5811, 9195, 3938, 5447,
                   6253, 3427, 8296, 5473, 3938, 6657),
  tn_load_kgyr = c(69120, 98990, 47100, 84780, 127280, 57890, 79860,
                   91580, 53420, 114800, 80020, 58920, 95280)
)

# Compute flow-weighted mean concentrations
# Concentration (µg/L) = load (kg/yr) / volume (m³/yr) * 1e6
hawqs_annual <- hawqs_annual |>
  dplyr::mutate(
    tp_conc_ugl = tp_load_kgyr / flow_m3yr * 1e6,
    tn_conc_ugl = tn_load_kgyr / flow_m3yr * 1e6
  )

cat("--- OK-HAWQS Annual Load Summary ---\n")
cat(sprintf("Mean annual flow:    %.1f million m³/yr\n",
            mean(hawqs_annual$flow_m3yr) / 1e6))
cat(sprintf("Mean annual TP load: %.0f kg/yr\n",
            mean(hawqs_annual$tp_load_kgyr)))
cat(sprintf("FWM TP conc:         %.1f µg/L\n",
            mean(hawqs_annual$tp_conc_ugl)))
cat(sprintf("FWM TN conc:         %.1f µg/L\n",
            mean(hawqs_annual$tn_conc_ugl)))
```

## Converting HAWQS Loads to okBATHTUB Inputs

The key conversion is from HAWQS load (kg/yr) and flow (m³/yr) to okBATHTUB inflow concentration (µg/L):

$$C_{inflow} (\mu g/L) = \frac{L (kg/yr) \times 10^6}{Q (m^3/yr)}$$

```{r hawqs_to_bathtub}
# Use long-term mean conditions for steady-state BATHTUB
mean_flow_m3yr <- mean(hawqs_annual$flow_m3yr)
mean_tp_ugl    <- mean(hawqs_annual$tp_conc_ugl)
mean_tn_ugl    <- mean(hawqs_annual$tn_conc_ugl)

cat(sprintf("BATHTUB inputs:\n"))
cat(sprintf("  Inflow volume: %.2f million m³/yr\n", mean_flow_m3yr / 1e6))
cat(sprintf("  TP inflow:     %.1f µg/L\n", mean_tp_ugl))
cat(sprintf("  TN inflow:     %.1f µg/L\n", mean_tn_ugl))
```

## Baseline BATHTUB Run

```{r baseline_run}
# Reservoir morphometry: Arcadia Lake (Cross Timbers ecoregion)
# surface_area_ha = 890, mean_depth_m = 4.2 (hardcoded for reproducibility)

baseline <- ok_load(
    inflow_m3yr   = mean_flow_m3yr,
    tp_inflow_ugl = mean_tp_ugl,
    tn_inflow_ugl = mean_tn_ugl,
    coefficients  = "oklahoma",
    ecoregion     = "Cross Timbers",
    segment_label = "baseline"
  ) |>
  ok_hydraulics(
    surface_area_ha = 890,
    mean_depth_m    = 4.2
  )

result_baseline <- baseline |>
  ok_retention() |>
  ok_inlake()    |>
  ok_tsi()

summary(result_baseline)
```

## BMP Scenario Analysis

OK-HAWQS/SWAT is designed to simulate the effect of conservation practices (cover crops, reduced tillage, buffer strips, wetland restoration) on watershed loading. Each HAWQS scenario produces a modified load — which feeds directly into okBATHTUB.

Here we demonstrate using simulated HAWQS BMP scenario outputs:

```{r bmp_scenarios}
# Simulated HAWQS BMP scenario outputs
# Each represents a different watershed management alternative
hawqs_scenarios <- list(
  list(
    label        = "Baseline (current conditions)",
    flow_m3yr    = mean_flow_m3yr,
    tp_load_kgyr = mean(hawqs_annual$tp_load_kgyr),
    tn_load_kgyr = mean(hawqs_annual$tn_load_kgyr)
  ),
  list(
    label        = "10% cropland cover crops",
    flow_m3yr    = mean_flow_m3yr * 0.97,     # slight flow reduction
    tp_load_kgyr = mean(hawqs_annual$tp_load_kgyr) * 0.88,
    tn_load_kgyr = mean(hawqs_annual$tn_load_kgyr) * 0.92
  ),
  list(
    label        = "30% cropland cover crops + buffer strips",
    flow_m3yr    = mean_flow_m3yr * 0.94,
    tp_load_kgyr = mean(hawqs_annual$tp_load_kgyr) * 0.72,
    tn_load_kgyr = mean(hawqs_annual$tn_load_kgyr) * 0.78
  ),
  list(
    label        = "Full BMP suite (TMDL alternative)",
    flow_m3yr    = mean_flow_m3yr * 0.91,
    tp_load_kgyr = mean(hawqs_annual$tp_load_kgyr) * 0.55,
    tn_load_kgyr = mean(hawqs_annual$tn_load_kgyr) * 0.62
  )
)

# Convert each HAWQS scenario to okBATHTUB concentrations and run pipeline
scenario_results <- lapply(hawqs_scenarios, function(sc) {
  tp_ugl <- sc$tp_load_kgyr / sc$flow_m3yr * 1e6
  tn_ugl <- sc$tn_load_kgyr / sc$flow_m3yr * 1e6

  r <- ok_load(
      inflow_m3yr   = sc$flow_m3yr,
      tp_inflow_ugl = tp_ugl,
      tn_inflow_ugl = tn_ugl,
      coefficients  = "oklahoma",
      ecoregion     = "Cross Timbers"
    ) |>
    ok_hydraulics(
      surface_area_ha = 890,
      mean_depth_m    = 4.2
    ) |>
    ok_retention() |>
    ok_inlake()    |>
    ok_tsi()

  data.frame(
    scenario         = sc$label,
    tp_inflow_ugl    = round(tp_ugl, 1),
    tp_reduction_pct = round(100 * (1 - tp_ugl / mean_tp_ugl), 1),
    tp_inlake_ugl    = round(r$data$tp_inlake_ugl, 1),
    chla_ugl         = round(r$data$chla_ugl,      2),
    secchi_m         = round(r$data$secchi_m,      2),
    tsi_mean         = round(r$data$tsi_mean,      1),
    trophic_state    = r$data$trophic_state,
    stringsAsFactors = FALSE
  )
})

scenario_df <- do.call(rbind, scenario_results)
print(scenario_df)
```

## Interannual Variability Analysis

Oklahoma's flood-dominated hydrology means that annual TP loading varies substantially between wet and dry years. A robust BATHTUB analysis should bracket this variability rather than relying on a single mean year:

```{r interannual}
# Run BATHTUB for each year in the HAWQS record
annual_results <- lapply(seq_len(nrow(hawqs_annual)), function(i) {
  yr   <- hawqs_annual[i, ]
  tp_ugl <- yr$tp_conc_ugl
  tn_ugl <- yr$tn_conc_ugl

  r <- ok_load(
      inflow_m3yr   = yr$flow_m3yr,
      tp_inflow_ugl = tp_ugl,
      tn_inflow_ugl = tn_ugl,
      coefficients  = "oklahoma",
      ecoregion     = "Cross Timbers"
    ) |>
    ok_hydraulics(
      surface_area_ha = 890,
      mean_depth_m    = 4.2
    ) |>
    ok_retention() |>
    ok_inlake()    |>
    ok_tsi()

  data.frame(
    year          = yr$year,
    flow_m3yr     = yr$flow_m3yr / 1e6,
    tp_inflow_ugl = round(tp_ugl, 1),
    tp_inlake_ugl = round(r$data$tp_inlake_ugl, 1),
    chla_ugl      = round(r$data$chla_ugl, 2),
    secchi_m      = round(r$data$secchi_m, 2),
    tsi_mean      = round(r$data$tsi_mean, 1),
    trophic_state = r$data$trophic_state
  )
}) |>
  do.call(rbind, args = _)

cat("--- Interannual Range ---\n")
cat(sprintf("TSI range:    %.1f - %.1f\n",
            min(annual_results$tsi_mean), max(annual_results$tsi_mean)))
cat(sprintf("Chl-a range:  %.2f - %.2f µg/L\n",
            min(annual_results$chla_ugl), max(annual_results$chla_ugl)))
cat(sprintf("Secchi range: %.2f - %.2f m\n",
            min(annual_results$secchi_m), max(annual_results$secchi_m)))
```

## Load-Response Curve with HAWQS Context

The load-response curve places the HAWQS scenarios in the context of the full TP range:

```{r response_plot, fig.height=5.5, eval=requireNamespace("ggplot2", quietly=TRUE)}
ok_plot_response(
  baseline,
  response     = "tsi",
  target_class = "mesotrophic",
  current_tp   = mean_tp_ugl,
  lake_name    = "Arcadia Lake -- OK-HAWQS/okBATHTUB Linkage"
)
```

## Best Practices for HAWQS-BATHTUB Linkage

**Temporal averaging** — BATHTUB is a steady-state model. Use the growing-season (May–October) mean for response variables and the water-year mean for loads. For interannual analysis, run BATHTUB separately for each year in the HAWQS record.

**Wet vs dry year stratification** — consider running separate BATHTUB analyses for wet-year (above median flow) and dry-year (below median flow) conditions from your HAWQS record to bracket uncertainty.

**Load vs concentration** — HAWQS produces loads (kg/yr). Convert to concentration (µg/L) using `C = L / Q * 1e6` before calling `ok_load()`. Verify that the resulting concentration is physically plausible for the watershed — very high concentrations with low flows may indicate numerical artefacts in the HAWQS calibration.

**Retention sensitivity** — BATHTUB's Larsen-Mercier retention is sensitive to hydraulic residence time, which depends on inflow volume. In drought years, lower inflow → longer residence time → higher retention → lower in-lake TP. Check that this behaviour is captured when comparing wet and dry year scenarios.

**Uncertainty propagation** — both HAWQS and BATHTUB have calibration uncertainty. Report predictions as ranges (e.g., ± one standard deviation across years) rather than single values, particularly for regulatory applications.

## References

- Soil and Water Assessment Tool (SWAT) documentation: https://swat.tamu.edu
- OK-HAWQS: Oklahoma application of HAWQS 2.0 (USGS/EPA national platform)
- Walker, W.W. (1996). *Simplified Procedures for Eutrophication Assessment and Prediction: User Manual*. U.S. Army Corps of Engineers, Instruction Report W-96-2.
- Oklahoma Water Resources Board (2024). FY-24 604(b) BIL Watershed Modeling Project — Sensitive Water Supply Reservoirs.
