---
title: "Oklahoma Reservoir Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Oklahoma Reservoir Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi       = 120
)
```

## Overview

This vignette demonstrates the full Oklahoma-specific workflow for reservoir eutrophication modelling with okBATHTUB. It covers:

1. Pulling in-lake water quality data from AWQMS via `ok_from_awqms()`
2. Running the pipeline with Oklahoma LMP-calibrated coefficients
3. Multi-segment reservoir modelling
4. Load reduction scenario analysis
5. Visualising results

All functions that connect to AWQMS require access to the OWRB network. Code blocks in this vignette are shown for illustration; set `eval = TRUE` in your own R session to run them.

## Setup

```{r load}
library(okBATHTUB)
library(dplyr)
```

## Step 1 — Pull Lake Monitoring Data from AWQMS

`ok_from_awqms()` connects to AWQMS, retrieves growing-season surface grab samples for the target lake(s), and returns a tidy annual summary data frame. It handles unit conversion, non-detect substitution, TN reconstruction, and phaeophytin-corrected chlorophyll prioritisation automatically.

```{r awqms_pull}
# Pull 10 years of data for Arcadia Lake
arcadia_data <- ok_from_awqms(
  lake_names            = "Arcadia Lake",
  year_start            = 2015,
  year_end              = 2024,
  ecoregion_lookup_path = ok_path("lake_ecoregion_lookup.csv")
)

glimpse(arcadia_data)
```

The output contains one row per monitoring station per year, with growing-season means for TP, TN, chlorophyll-a (phaeophytin-corrected), Secchi depth, TSS, and turbidity, plus ecoregion and sample counts for each parameter.

### Observed trophic state from monitoring data

Before modelling, compute observed TSI from the monitoring data to establish a baseline:

```{r observed_tsi}
obs_tsi <- arcadia_data |>
  dplyr::filter(!is.na(tp_ugl), !is.na(chla_ugl), !is.na(secchi_m)) |>
  rowwise() |>
  dplyr::mutate(
    tsi_tp     = 14.42 * log(tp_ugl)   + 4.15,
    tsi_chla   = 9.81  * log(chla_ugl) + 30.6,
    tsi_secchi = 60.0  - 14.41 * log(secchi_m),
    tsi_mean   = mean(c(tsi_tp, tsi_chla, tsi_secchi))
  ) |>
  ungroup()

# Annual mean TSI by site
obs_tsi |>
  dplyr::group_by(sample_year) |>
  dplyr::summarise(
    tsi_tp     = round(mean(tsi_tp,     na.rm = TRUE), 1),
    tsi_chla   = round(mean(tsi_chla,   na.rm = TRUE), 1),
    tsi_secchi = round(mean(tsi_secchi, na.rm = TRUE), 1),
    tsi_mean   = round(mean(tsi_mean,   na.rm = TRUE), 1),
    .groups    = "drop"
  ) |>
  print()
```

## Step 2 — Single-Segment Pipeline with Oklahoma Coefficients

Use the reservoir morphometry lookup and Oklahoma ecoregion-specific coefficients:

```{r single_segment}
# Look up reservoir morphometry
res <- ok_reservoir("Arcadia Lake")
cat("Surface area:", "Arcadia Lake", "ha | Mean depth:", "Arcadia Lake", "m\n")
cat("Ecoregion:   ", "Arcadia Lake", "\n")

# Estimated inflow TP from watershed monitoring
# (Replace with USGS NWIS/FLUX-derived annual load estimate)
tp_inflow_est  <- 120   # µg/L flow-weighted mean
tn_inflow_est  <- 1800  # µg/L
inflow_vol_est <- 45e6  # m³/yr (~1.43 m³/s mean)

result <- ok_load(
    inflow_m3yr   = inflow_vol_est,
    tp_inflow_ugl = tp_inflow_est,
    tn_inflow_ugl = tn_inflow_est,
    coefficients  = "oklahoma",
    ecoregion     = "Cross Timbers"
  ) |>
  ok_hydraulics(
    surface_area_ha = 890,
    mean_depth_m    = 4.2
  ) |>
  ok_retention() |>
  ok_inlake()    |>
  ok_tsi()

summary(result)
```

### Compare predicted vs observed

```{r compare}
cat("--- Predicted (okBATHTUB) vs Observed (LMP) ---\n")
cat(sprintf("In-lake TP:   predicted = %.1f µg/L | observed = %.1f µg/L\n",
            result$data$tp_inlake_ugl,
            mean(arcadia_data$tp_ugl, na.rm = TRUE)))
cat(sprintf("Chl-a:        predicted = %.2f µg/L | observed = %.2f µg/L\n",
            result$data$chla_ugl,
            mean(arcadia_data$chla_ugl, na.rm = TRUE)))
cat(sprintf("Secchi depth: predicted = %.2f m    | observed = %.2f m\n",
            result$data$secchi_m,
            mean(arcadia_data$secchi_m, na.rm = TRUE)))
```

## Step 3 — Multi-Segment Reservoir

For reservoirs with pronounced longitudinal gradients (common in Oklahoma's flood-dominated, elongated impoundments), the multi-segment approach produces more accurate predictions than treating the whole reservoir as a single mixed pool.

Eufaula Lake — Oklahoma's largest reservoir — has a strongly stratified riverine-to-lacustrine gradient:

```{r multi_segment}
eufaula <- ok_reservoir("Eufaula Lake")

# Three segments representing the longitudinal gradient
# Segment proportions estimated from USACE bathymetric data
segments <- list(
  list(
    label           = "Riverine",
    surface_area_ha = eufaula$surface_area_ha * 0.25,
    mean_depth_m    = 3.2
  ),
  list(
    label           = "Transitional",
    surface_area_ha = eufaula$surface_area_ha * 0.35,
    mean_depth_m    = 4.8
  ),
  list(
    label           = "Lacustrine",
    surface_area_ha = eufaula$surface_area_ha * 0.40,
    mean_depth_m    = 7.1
  )
)

chain <- ok_segment_chain(
  inflow_m3yr   = 2200e6,   # Eufaula: large watershed, high inflow
  tp_inflow_ugl = 145,
  tn_inflow_ugl = 2100,
  segments      = segments,
  coefficients  = "oklahoma",
  ecoregion     = "Arkansas Valley"   # statewide pooled applied (low R2)
)

ok_segment_summary(chain) |>
  dplyr::select(segment, tp_inflow_ugl, tp_inlake_ugl, tp_retention,
                chla_ugl, secchi_m, tsi_mean, trophic_state) |>
  print()
```

### Longitudinal profile plot

```{r segment_plot, fig.height=5.5, eval=requireNamespace("ggplot2", quietly=TRUE) && exists("chain")}
ok_plot_segments(chain, lake_name = "Eufaula Lake")
```

The longitudinal profile illustrates the transition from a nutrient-rich, turbid riverine zone to a more dilute lacustrine pool — a pattern typical of Oklahoma's large flood-control reservoirs.

## Step 4 — Load Reduction Scenario Analysis

The scenario analysis workflow answers: *how much does inflow TP need to be reduced to achieve a water quality target?*

```{r scenarios}
baseline <- ok_load(
    inflow_m3yr   = inflow_vol_est,
    tp_inflow_ugl = tp_inflow_est,
    tn_inflow_ugl = tn_inflow_est,
    coefficients  = "oklahoma",
    ecoregion     = "Cross Timbers"
  ) |>
  ok_hydraulics(
    surface_area_ha = 890,
    mean_depth_m    = 4.2
  )

# Sweep 10% through 70% TP reduction, target: mesotrophic (TSI < 50)
sweep <- ok_scenario_sweep(
  baseline,
  max_reduction_pct = 70,
  step_pct          = 10,
  target_class      = "mesotrophic"
)

sweep |>
  dplyr::select(scenario, tp_inflow_ugl, tp_inlake_ugl,
         chla_ugl, secchi_m, tsi_mean, trophic_state, meets_target) |>
  print()
```

### Find the minimum reduction needed

```{r min_reduction}
min_met <- sweep |>
  dplyr::filter(meets_target == TRUE) |>
  dplyr::slice(1)

if (nrow(min_met) > 0) {
  cat(sprintf(
    "Minimum TP reduction to achieve mesotrophic (TSI < 50): %s\n",
    min_met$scenario
  ))
  cat(sprintf(
    "  Inflow TP: %.0f µg/L → In-lake TP: %.1f µg/L → TSI: %.1f\n",
    min_met$tp_inflow_ugl, min_met$tp_inlake_ugl, min_met$tsi_mean
  ))
} else {
  cat("No scenario in the sweep achieves mesotrophic status.\n")
  cat("Consider evaluating reductions beyond", max(sweep$tp_reduction_pct, na.rm=TRUE), "%\n")
}
```

### Custom scenarios

`ok_scenario()` accepts any combination of TP reduction, TN reduction, or flow change:

```{r custom_scenarios}
custom <- ok_scenario(
  baseline  = baseline,
  scenarios = list(
    list(label = "Agricultural BMPs (25% TP)",
         tp_reduction = 0.25),
    list(label = "Wastewater upgrade (40% TP, 30% TN)",
         tp_reduction = 0.40, tn_reduction = 0.30),
    list(label = "Drought year (20% flow reduction)",
         flow_change  = -0.20),
    list(label = "Wet year + BMPs",
         tp_reduction = 0.25, flow_change = 0.15),
    list(label = "TMDL target TP = 60 µg/L",
         tp_inflow_ugl = 60)
  ),
  target_class = "mesotrophic"
)

custom |>
  dplyr::select(scenario, tp_inflow_ugl, tsi_mean, trophic_state, meets_target) |>
  print()
```

## Step 5 — Visualisation

### Load-response curve

```{r response_plot, fig.height=5.5, eval=requireNamespace("ggplot2", quietly=TRUE) && exists("baseline")}
ok_plot_response(
  baseline,
  response     = "tsi",
  target_class = "mesotrophic",
  current_tp   = tp_inflow_est,
  lake_name    = "Arcadia Lake"
)
```

### Scenario comparison

```{r scenario_plot, fig.height=5, eval=requireNamespace("ggplot2", quietly=TRUE) && exists("sweep")}
ok_plot_scenario(sweep, lake_name = "Arcadia Lake")
```

### Carlson TSI diagram from observed data

The TSI deviation diagram is a standard diagnostic for separating algal limitation from non-algal turbidity:

```{r tsi_diagram, eval=FALSE}
# Pull all LMP lakes for statewide TSI diagram
all_lakes <- ok_from_awqms(
  year_start            = 2018,
  year_end              = 2024,
  ecoregion_lookup_path = ok_path("lake_ecoregion_lookup.csv")
)

# Compute observed TSI
all_tsi <- all_lakes |>
  dplyr::filter(!is.na(tp_ugl), !is.na(chla_ugl)) |>
  dplyr::mutate(
    tsi_tp   = 14.42 * log(tp_ugl)   + 4.15,
    tsi_chla = 9.81  * log(chla_ugl) + 30.6
  )

ok_plot_tsi(all_tsi, color_by = "eco_l3_name")
```

Points below the 1:1 diagonal (TSI(Chl-a) < TSI(TP)) indicate phosphorus-rich lakes where algal biomass is suppressed relative to nutrient levels — the signature of light limitation by inorganic turbidity common in Oklahoma's western and Arkansas Valley reservoirs.

## Workflow Summary

The complete Oklahoma reservoir workflow in one code block:

```{r full_workflow}
# 1. Look up reservoir morphometry
res <- ok_reservoir("Arcadia Lake")

# 2. Define loading inputs (from USGS/FLUX load estimates)
loading <- list(
  inflow_m3yr   = 45e6,
  tp_inflow_ugl = 120,
  tn_inflow_ugl = 1800
)

# 3. Run pipeline with Oklahoma coefficients
baseline <- ok_load(
    inflow_m3yr   = loading$inflow_m3yr,
    tp_inflow_ugl = loading$tp_inflow_ugl,
    tn_inflow_ugl = loading$tn_inflow_ugl,
    coefficients  = "oklahoma",
    ecoregion     = "Cross Timbers"
  ) |>
  ok_hydraulics(
    surface_area_ha = 890,
    mean_depth_m    = 4.2
  )

result <- baseline |>
  ok_retention() |>
  ok_inlake()    |>
  ok_tsi()

# 4. Scenario analysis
sweep <- ok_scenario_sweep(baseline, max_reduction_pct = 60,
                            step_pct = 10, target_class = "mesotrophic")

# 5. Visualise
ok_plot_response(baseline, response = "tsi",
                  target_class = "mesotrophic",
                  current_tp = loading$tp_inflow_ugl,
                  lake_name = "Arcadia Lake")
```

## References

- Walker, W.W. (1996). *Simplified Procedures for Eutrophication Assessment and Prediction: User Manual*. U.S. Army Corps of Engineers, Instruction Report W-96-2.
- Oklahoma Water Resources Board (2024). Lake Monitoring Program data. AWQMS database. Oklahoma City, OK.
- Jones, J.R. and Knowlton, M.F. (2005). Chlorophyll response to nutrients and non-algal seston in Missouri reservoirs. *Lake and Reservoir Management*, 21(3), 361–371.
- Dzialowski, A.R. et al. (2011). Effects of non-algal turbidity on cyanobacterial biomass in seven turbid Kansas reservoirs. *Lake and Reservoir Management*, 27(1), 6–14.
